Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-12 paper.tex
Rephrase

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because our government-approved programming language does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete datastructure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. exploit parametricity for
reasoning.

@@

The cube drawn in Figure~\ref{fig:state} could just be modeled as
%
\coqe!Rows.T (Cols.T (Slice8.T bool))!,
%
\coqe!Slice8.T (Rows.T (Cols.T bool))!,
%
or any other composition of these 3 type constructors. They are all
isomorphic to a sequence of 128 bits. We shall wait until
Section~\ref{sec:assembling} to settle on the most natural
representation for our specification effort. We revisit this choice in
Section~\ref{sec:bitslicing} when it comes to producing a
memory-efficient representation.

@@

Having modeled the core data type, we now turn to modeling the cipher
as a purely functional program. To stay clear from the temptation of
array mismanagement, we will forbid ourselves from index arithmetic:
no indexing an array-like structure by \coqe!i+1! or \coqe!j-i! in
this paper!

@@

Much like $\lambda$ in functional programming, the exclusive-or
features prominently in the cryptographic liturgy. In particular, it
is instrumental to mix the derived keys during the ciphertext
computation. For genericity, we model this layer as an operation
defined over any Boolean algebra~\citep{givant:boolean-algebras}:
%
\coqfrom{Skinny/Base}{addroundkey}

@@

The role of the diffusion layer is to divert individual bits across the
whole structure. Inevitably, this calls for a notion of indexing.
Naperian functors~\citep{hancock:napierian} identify a type \coqe!Ix! of
indices as the logarithm of an exponential type:
%
\coqfrom{Algebra}{Naperian}
%
through the fact that \coqe!lookup! and \coqe!init! form a bijection.
This is also known as a representable functor in mathematical
circles~\citep{maclane:working-categorist}

@@

Following our earlier discussion, we ask that \coqe!Rows.T!
(respectively, \coqe!Cols.T!) is a Naperian functor indexed by a set
of 4 elements. Similarly, \coqe!Slice8.T! is a Naperian functor
indexed by 8 elements. We defined
%
\coqfrom{Rows}{Ix}
%
as the logarithm of \coqe!Rows.T!,
%
\coqfrom{Cols}{Ix}
%
as the logarithm of \coqe!Cols.T!, and
%
\coqfrom{Slice8}{Ix}
%
as the logarithm of \coqe!Slice8.T!.

Note that we have a notion of indexing but no arithmetic: we remain in
line with our objectives. Note also that being Naperian implies being
a commutative applicative functor. As a corollary, we get that any
combination of \coqe!Rows.T!, \coqe!Cols.T! and \coqe!Slice8.t!
applied to \coqe!bool! yields a Boolean algebra (over which we can
therefore write combinational circuits).

@@

Note that, since they actually have the same number of elements, we
could also go the other way around, defining an inhabitant of
\coqe!Cols.T Rows.Ix!. We do not need this construction for \Skinny{}. It
is necessary in the case of \Gift{}, which proceeds by transposition of
$4 \times 4$ matrix.

@@

The first step of diffusion process consists in performing a right
rotation by a specific amount over each individual column. However, we
do not have enough structure to identify a ``right'' or ``left''
direction over our containers. The standard tool of the trade to
achieve this effect is the notion of foldable
functor~\citep{yorgey:typeclassopedia}:
%
\coqfrom{Algebra}{Foldable}

@@

After suitable generalization, this leads us to the following model
for the operation depicted in Figure~\ref{fig:shiftrows}
%
\coqfrom{Skinny/Base}{shiftrows}
%
which represents the first half of the diffusion layer.

%% * Application: mixcolumns
%% ** figure: psi_ >-> ror 1

The second half corresponds to a combinational circuit \coqe!psi_!
%
\coqfrom{Skinny/Base}{psi}
%
which exercises the Naperian structure of \coqe!Rows.T! together with
some underlying Boolean algebra. This is followed by a right rotation
of rows. Altogether, this defines the operation depicted in Figure~\ref{fig:mixcolumns}
%
\coqfrom{Skinny/Base}{mixcolumns}

>>>>

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because our government-approved programming language does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete data structure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. exploit parametricity for
reasoning.

@@

The cube drawn in Figure~\ref{fig:state} could just be modeled as
%
\coqe!Rows.T (Cols.T (Slice8.T bool))!,
%
\coqe!Slice8.T (Rows.T (Cols.T bool))!,
%
or any other composition of these 3 type constructors. They are all
isomorphic to a sequence of 128 bits. We shall wait until
Section~\ref{sec:assembling} to settle on the most natural
representation for our specification effort. We revisit this choice in
Section~\ref{sec:bitslicing} when we are concerned with producing a
memory-efficient representation.

@@

Having modeled the core data types, we now turn to modeling the cipher
as a purely functional program. To stay clear from the temptation of
array mismanagement, we disallow index arithmetic altogether: no
indexing an array-like structure by \coqe!i+1! or, even worse,
\coqe!j-i! in this paper!

@@

Much like $\lambda$ in functional programming, the exclusive-or
$\oplus$ features prominently in the cryptographic liturgy. In
particular, it is instrumental to mix the derived keys during
ciphertext computation. For genericity, we model this layer as an
operation defined over any Boolean
algebra~\citep{givant:boolean-algebras}:
%
\coqfrom{Skinny/Base}{addroundkey}

@@

The role of the diffusion layer is to divert individual bits across the
whole structure. Inevitably, this calls for a notion of indexing.
Naperian functors~\citep{hancock:napierian} identify a type \coqe!Ix! of
indices as the logarithm of an exponential type:
%
\coqfrom{Algebra}{Naperian}
%
through the fact that \coqe!lookup! and \coqe!init! form a bijection.
This is also known as a representable functor in categorical
circles~\citep{maclane:working-categorist}.

@@

Following our earlier discussion, we ask for \coqe!Rows.T!
(respectively, \coqe!Cols.T!) to be a Naperian functor indexed by a
set of 4 elements. Similarly, \coqe!Slice8.T! must be a Naperian
functor indexed by 8 elements. We define
%
\coqfrom{Rows}{Ix}
%
as the logarithm of \coqe!Rows.T!,
%
\coqfrom{Cols}{Ix}
%
as the logarithm of \coqe!Cols.T!, and
%
\coqfrom{Slice8}{Ix}
%
as the logarithm of \coqe!Slice8.T!.

Note that we have a notion of indexing but no arithmetic: we remain in
line with our objectives. Note also that being Naperian implies being
a commutative applicative functor. As a corollary, we get that any
combination of \coqe!Rows.T!, \coqe!Cols.T! and \coqe!Slice8.t!
applied to \coqe!bool! yields a Boolean algebra, over which we can
therefore write combinational circuits.

@@

Note that, since they actually have the same number of elements, we
could also go the other way around, defining an inhabitant of
\coqe!Cols.T Rows.Ix!. We do not need this construction for \Skinny{}
but it is necessary for \Gift{}, which proceeds by transposition of a
$4 \times 4$ matrix.

@@

The first step of diffusion process consists in applying a right
rotation over each individual column. However, we do not have enough
structure to identify a ``right'' or ``left'' direction over our
containers. The standard tool of the trade to achieve this effect is
the notion of foldable functor~\citep{yorgey:typeclassopedia}:
%
\coqfrom{Algebra}{Foldable}

@@

After suitable generalization, this leads us to the following model
for the \coqe!shiftrows_! operation, depicted in
Figure~\ref{fig:shiftrows}:
%
\coqfrom{Skinny/Base}{shiftrows}
%
which represents the first half of the diffusion layer.
%
%% * Application: mixcolumns
%% ** figure: psi_ >-> ror 1
%
The second half corresponds to a combinational circuit \coqe!psi_!
%
\coqfrom{Skinny/Base}{psi}
%
which exercises the Naperian structure of \coqe!Rows.T! together with
the underlying Boolean algebra. This is followed by a right rotation
of rows. Altogether, this defines the \coqe!mixcolumns_! operation
depicted in Figure~\ref{fig:mixcolumns}
%
\coqfrom{Skinny/Base}{mixcolumns}
